---
title: "APD report"
author: "Lewis Campbell"
date: "`r Sys.Date()`"
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
subtitle: Template for data reading and cleaning from the ANZICS APD
bibliography: null
link-citations: yes
params:
   year: 2021
   data: APD_2021.csv
   output-file: !r paste0("APD-", Sys.Date(), ".pdf")
---

```{r setup, include=FALSE, warning=FALSE}
library(tufte)
library(knitr)
library(httr)
library(tidyverse)
library(readxl)
library(lubridate)
library(ggnewscale)
library(gridExtra)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```
```{r download, echo=FALSE, message=FALSE}
# There is one way to get the data and one only. I can't programmatically access the data for now. May be https://comet.anzics.com.au/api/v1/reports/runreport which is the content of POST on developer tools in Firefox

#res <- POST(
#    url = "https://comet.anzics.com.au/#/login",
#    encode = "json",
#    body = list(
#        type = "ANZICS.Domain.Comet.Reporting.ReportDefinition, ANZICS.Domain",
#        careUnits = "57e139c6-cbec-e611-80c8-005056a105bd",
#        icuFromAdmissionDateTime = "2021-01-01T00:00:00+11:00",
#        icuToAdmissionDateTime = "2021-01-30T00:00:00+11:00",
#        report = "1"
#    )
#)


```


```{r cleaning, echo=FALSE, message=FALSE, warning=FALSE}
# Having the download as an Excel file is painful, because of lots of things. The worst is Excel's habit of turning various columns into character columns, silently; such as all the numeric columns here. Exporting (Save As...) from Excel to .csv makes some of these go away, but introduces some other problems, again because Excel has got up to some funny japes behind the scenes. The column types pretty much all have to be exhaustively specified, including the eccentrically named Trachestomy. If one column parses properly in csv and Excel it can be left out, but just...don't.

# First go to https://comet.anzics.com.au, log in and select the download for the period you require. Save it, then open it. This time "Save as ..." and ensure the type is .csv (and if it asks, save with UTF-8 encoding, never mind what that means for now). Plug whatever you call it into APD_source below.

APD_source <- paste0("./APD_", params$year, ".csv")

cols_APD <- cols(
   APD_Inclusion = col_character(),
   ANZROD_Inclusion = col_character(),
   APACHE3_Inclusion = col_character(),
   `HRN/NIH` = col_character(),
   PatientId = col_character(),
   FirstName = col_character(),
   LastName = col_character(),
   Address = col_character(),
   Suburb = col_character(),
   State = col_character(),
   Country = col_character(),
   Postcode = col_character(),
   SLK = col_character(),
   AgeICU = col_double(),
   AgeHosp = col_double(),
   DOB = col_date(format = "%d/%m/%Y"),
   Sex = col_character(),
   Indigenous = col_character(),
   Ethnicity = col_character(),
   EthnicityOther = col_logical(),
   Height = col_double(),
   Weight = col_double(),
   CareUnitAdmID = col_character(),
   CareUnit = col_character(),
   CareType = col_character(),
   YearAdm = col_double(),
   FYAdm = col_double(),
   MonthYearAdm = col_integer(),
   HOSP_ADM_DTM = col_datetime(format = "%d/%m/%Y %H:%M:%S"),
   HOSP_DIS_DTM = col_datetime(format = "%d/%m/%Y %H:%M:%S"),
   ICU_ADM_DTM = col_datetime(format = "%d/%m/%Y %H:%M:%S"),
   ICU_DIS_DTM = col_datetime(format = "%d/%m/%Y %H:%M:%S"),
   ICU_DIS_DEC_DTM = col_datetime(format = "%d/%m/%Y %H:%M:%S"),
   ICU_ADM_DT = col_date(format = "%d/%m/%Y"),
   ICU_ADM_TM = col_time(format = ""),
   ICU_DS_DT = col_date(format = "%d/%m/%Y"),
   ICU_DS_TM = col_time(format = ""),
   ReadmissionLagHours = col_double(),
   ICUadm_count = col_double(),
   ICUSource = col_character(),
   ICULOS = col_double(),
   LongStay_7 = col_character(),
   LongStay_14 = col_character(),
   ICUOutcome = col_character(),
   DiedICU = col_character(),
   HOSLOS = col_double(),
   HospitalSource = col_character(),
   TransferredFrom = col_character(),
   HospitalOutcome = col_character(),
   TransferredTo = col_character(),
   DiedHospital = col_character(),
   DataComplete = col_character(),
   APACHEDiagnosis = col_double(),
   APACHEDiagText = col_character(),
   APACHESubcode = col_double(),
   APACHECategory = col_character(),
   Med_Surg = col_character(),
   AP2score = col_double(),
   AP3score = col_double(),
   AP3ROD = col_double(),
   ANZROD = col_double(),
   ICUAdmission = col_character(),
   ReadmissionEpisode = col_double(),
   ElectiveSurg = col_character(),
   PlannedAdm = col_character(),
   METAdm = col_character(),
   PregStatus = col_character(),
   ThromboPro = col_character(),
   TreatmentGoals = col_character(),
   SmokingStatus = col_character(),
   AfterhoursDis = col_character(),
   ExitBlockHrs = col_double(),
   GCSDateTime = col_datetime(format = "%d/%m/%Y %H:%M:%S"),
   GCS = col_integer(),
   GCSEye = col_integer(),
   GCSVerb = col_integer(),
   GCSMotor = col_integer(),
   GCSSedated = col_integer(),
   CardArrest = col_character(),
   Diabetes = col_character(),
   Frailty = col_character(),
   Delirium = col_character(),
   PressureInjury = col_character(),
   ARF = col_character(),
   ChronicRespiratory = col_character(),
   ChronicCardiovascular = col_character(),
   Cirrhosis = col_character(),
   ChronicRenal = col_character(),
   ImmuneDisease = col_character(),
   ImmunoSuppressed = col_character(),
   AIDS = col_character(),
   HepaticFailure = col_character(),
   Lymphoma = col_character(),
   MetastaticCancer = col_character(),
   Leukaemia = col_character(),
   CABGGraft = col_logical(),
   CABGRedo = col_logical(),
   AMIThromboTher = col_character(),
   ECMO = col_character(),
   Inotropes = col_character(),
   InvVent = col_character(),
   NIVVent = col_character(),
   RenalRep = col_character(),
   Trachestomy = col_character(),
   INVDayOne = col_character(),
   INV_Hrs = col_double(),
   NIV_Hrs = col_double(),
   HFNC_Hrs = col_double(),
   PT_identifier = col_logical(),
   HOSP_identifier = col_character(),
   ICU_identifier = col_logical(),
   ALBUMHI = col_double(),
   ALBUMLO = col_double(),
   AP3FIO = col_double(),
   AP3PO2 = col_double(),
   AP3CO2O = col_double(),
   AP3PH = col_double(),
   Intubated = col_character(),
   BILI = col_double(),
   CREATHI = col_double(),
   CREATLO = col_double(),
   DIASTOLICHI = col_double(),
   DIASTOLICLO = col_double(),
   FIO2 = col_double(),
   PAO2 = col_double(),
   PACO2 = col_double(),
   PH = col_double(),
   GLUCHI = col_double(),
   GLUCLO = col_double(),
   HCO3HI = col_double(),
   HCO3LO = col_double(),
   HCTHI = col_double(),
   HCTLO = col_double(),
   HMGNHI = col_double(),
   HMGNLO = col_double(),
   HRHI = col_double(),
   HRLO = col_double(),
   KHI = col_double(),
   KLO = col_double(),
   LACTATE = col_double(),
   MAPHI = col_double(),
   MAPLO = col_double(),
   NAHI = col_double(),
   NALO = col_double(),
   RRHI = col_double(),
   RRHI_VENT = col_character(),
   RRLO = col_double(),
   RRLO_VENT = col_character(),
   SYSTOLICHI = col_double(),
   SYSTOLICLO = col_double(),
   TEMPHI = col_double(),
   TEMPLO = col_double(),
   PLATHI = col_double(),
   PLATLO = col_double(),
   UREA = col_double(),
   WCCHI = col_double(),
   WCCLO = col_double(),
   URINEOP = col_double(),
   FreeVoiding = col_character(),
   PatientEntityId = col_character(),
   HospitalAdmissionEntityId = col_character()
   )

# Ctrl+C Copy the folder path for the csv file in the directory then use this:
# writeClipboard(gsub("\\\\", "/", readClipboard()))
# Then paste the clipboard contents into the "file" entry in the call to read_csv below
# read it in as the object called APD, which will appear in the Data window
APD <- read_csv(file = APD_source, na = c("", "NA", "NULL"), col_types = cols_APD)
# Here is the behaviour of the Excel option without specifying types. Expect all numerical values to be classed as Character, and the col_types argument doesn't behave the same in readxl:: (wants a character) as in readr:: (wants a list)

#APD_excel_test <- read_xlsx("C:/Users/lewis/Documents/The_Information/Research_of_me/Observational/APD_Report/COMET-Extract-APD.xlsx", sheet = 1, na = c("", "NA", "NULL"), n_max = 1060)
#str(APD_excel_test)

# and check for the numbers of certain diagnoses: head injury with
#sum(APD$APACHEDiagnosis %in% c("601" , "1601"))
# The use of Yes and No for logical vectors. Hard to fix these. The basic command is recode(illogical_string, `Yes` = TRUE, `No` = FALSE). This command uses a pipe %>% to send the object to the next function, then mutate() to change the content of a column. It's scoped with across(), to mutate across all the named columns, 

APD <- APD %>% 
   mutate(across(c(Indigenous, ElectiveSurg:METAdm, AfterhoursDis, Delirium:Leukaemia, ECMO:INVDayOne, CardArrest, Intubated, FreeVoiding), ~recode(., `Yes` = TRUE, `No` = FALSE, `Contraindicated` = TRUE, `NotIndicated` = TRUE, `Died` = TRUE, `Survived` = FALSE, .default = NA))) %>% mutate(PF = AP3PO2/AP3FIO)

# Manipulates the data for the period directly. Use the commented-out filter (by removing the hash before the %>% signs) to return only the ones where there were any actual deaths.

APDmortality <- APD %>% 
   select(PatientId, APACHECategory, ANZROD, ICUOutcome) %>% 
   mutate(Mortality = as.numeric(ICUOutcome=="Died")) %>% 
   group_by("Category" = APACHECategory) %>%  
   summarise("Admissions" = n(), 
             "Observed Mortality" = round(mean(Mortality, na.rm=T), 3), 
             "ANZROD Predicted Mortality" = round(mean(ANZROD, na.rm=T), 3)) 

# Regenerates out of hours discharge from within the data 

APD <- APD %>%
   mutate(OOHDC = case_when(
      hour(ICU_DIS_DTM)>18 | hour(ICU_DIS_DTM)<6 ~ "After_hours",
      day(ICU_DIS_DTM) %in% c(1,7) ~ "Weekend",
      TRUE ~ "In_hours"
      )) %>% 
   mutate(ExitBlockHrs = time_length(ICU_DIS_DTM - ICU_DIS_DEC_DTM, unit = "hour")) %>% 
   mutate(missed_sepsis = ICU_ADM_DTM-HOSP_ADM_DTM>172800 & APACHEDiagnosis %in% c(501:504, 212)) %>% 
   mutate(nointervention = Inotropes ==FALSE & InvVent ==FALSE & INVDayOne == FALSE & NIVVent == FALSE & RenalRep ==FALSE & Trachestomy == FALSE)

```

# Introduction

This report used the input file `r APD_source`, covering the period `r min(APD$ICU_ADM_DT)` to `r max(APD$ICU_DS_DT, na.rm=TRUE)`. 

The ANZICS APD is a registry covering every adult patient admitted to the majority of Intensive Care Units in Australia^[The explanation and supporting materials are available from [ANZICS](https://www.anzics.com.au/anzics-registries).]. We contributed to the benchmarking of illness and treatment in the [Northern Territory](https://www.nt.gov.au) by entering every patient into this database. Using the statistical programming environment `R` with the `RStudio` Integrated Development Environment, we are able to perform sophisticated analyses. We start with a [recurring report](#regular) containing requested items in a common format introduction describing the total patient cohort over the period. 

After that we have a section laying out the evidence for [another question](#question). In that section there is a little more indication of the `R` code which generated that evidence, for transparency. The entire code is available on [github](https://github.com/lewisfromlewis/APD_report), with a brief synthetic dataset to explore the effect of new code without access to the original data.

# Report {#regular}

## Numbers of patients 
Over the period there were `r length(APD$SLK)` admissions, of `r length(unique(APD$SLK))` individuals. The weekly admission rate is shown below, and the instantaneous occupancy, changing at the moment of each admission or discharge, is in the margin.

```{r admissions, echo=FALSE, message=FALSE, warning=FALSE}
admission_rate <- ggplot(APD, mapping = aes(x=ICU_ADM_DTM))

admission_rate + geom_histogram(fill = "pink", alpha = 0.9, show.legend = FALSE)+
   labs(title = "Admission rate over the period",
        x = "Admission date",
        y = "Admissions per week")
   
```

```{r instantaneous.occupancy, echo = FALSE, message = FALSE, warning = FALSE}
# Writes a data frame containing an ordered list of times, with the type of each time, and a running total which is +1 for each admission time and -1 for each discharge time
Patient_movements <- APD %>% 
   select(PatientId, Admit = ICU_ADM_DTM, Discharge = ICU_DIS_DTM, ANZROD = ANZROD) %>% 
   gather(key = "Event", value = "Time", -PatientId, -ANZROD) %>% 
   arrange(Time)  
# Cumulative occupancy, changes instantaneously with every admission or discharge event 
cum_occ <- rep(0, length(Patient_movements$Event)); for(i in seq_along(Patient_movements$Event)){
   if(i == 1)
   {cum_occ <- 1}
   else if(Patient_movements$Event[i] == "Admit")
      {cum_occ[i] = cum_occ[i-1]+1}
   else if(Patient_movements$Event[i] == "Discharge")
   {cum_occ[i] = cum_occ[i-1]-1} 
}

# Cumulative Intensity Weighted Occupancy, a simple (and simplistic?) transformation multiplying the value of each event by the _admission_ ANZROD
IWtO <- rep(0, length(Patient_movements$Event)); for(i in seq_along(Patient_movements$Event)){
   if(i == 1)
      {IWtO <- Patient_movements$ANZROD[i]}
   else if(is.na(Patient_movements$ANZROD[i]))
      {IWtO[i] <- IWtO[i-1]}
   else if(Patient_movements$Event[i] == "Admit")
      {IWtO[i] = IWtO[i-1]+Patient_movements$ANZROD[i]}
   else if(Patient_movements$Event[i] == "Discharge")
   {IWtO[i] = IWtO[i-1]-Patient_movements$ANZROD[i]} 
}

Patient_movements <- Patient_movements %>% 
   mutate(Occupancy = cum_occ, IWO = IWtO)

Moves_plot <- ggplot(Patient_movements, mapping = aes(x=Time, y=Occupancy))

Weight_moves_plot <- ggplot(Patient_movements, mapping = aes(x = Time))

```

```{r Moves.plot, fig.margin = TRUE, fig.cap = "Instantaneous occupancy over the period.", fig.width=3.5, fig.height=3.5, cache=TRUE, message=FALSE, warning=FALSE, echo = FALSE}
Moves_plot + 
   geom_line(colour = "darkgray") +
   theme(panel.background = element_blank()) +
   labs(title = "Instantaneous Occupancy",
        y = "Occupancy") 
```

## Intensity weighted occupancy 

`r newthought('The admission rate tells part of the story')`, but sicker patients need more intense care. The severity score on admission tells a little more of the story. Each bar is an individual patient admission. The vertical height of the bars is equal to the predicted risk of death on admission, and the horizontal width is the length of the admission. Transparency is set at 0.1 to allow overlapping bars to be seen. This gives a good qualitative view of how many very sick patients were admitted at a time, or "Intensity weighted bed occupancy". The rest of the story about treatment intensity can only be told with daily data, which we don't compile. 

```{r Admission-severity, fig.fullwidth = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Patients sequentially represented on the y axis with the length of bar indicating length of stay

occupancy_intensity <- ggplot(data = APD)

occupancy_intensity +
   geom_rect(mapping = aes(xmin = ICU_ADM_DTM, 
                           xmax = ICU_DIS_DTM, 
                           ymin = rank(ICU_ADM_DTM), 
                           ymax = rank(ICU_ADM_DTM)+(40*ANZROD), 
                           fill= ANZROD,
                           alpha = 0.1)) +
   labs(title = "Individual severity plot",
        subtitle = "Height and colour both indicate severity",
        x = "Length of bar is time spent in ICU",
        y = "Admission number",
        fill = "Severity on admission",
        alpha = "Transparency")
```
The gaps and overlaps give an idea of when it was busy, and the height and brightness of the bars give an idea of how sick each patient was on admission. 

Now the occupancy can be numerically weighted for the admission ANZROD score. This is very simplistic, but gives a granular summary of numbers combined with severity. In the plots below, firstly the occupancy is weighted for the severity of illness of the patients who are occupying the beds. The resulting intensity-weighted occupancy is then plotted over the raw occupancy, balanced so they display the same mean value. Then a locally weighted estimate is applied to show the deterministic portion of the trend over time. In simple language, it smooths the instantaneous data to draw the eye to trends. 

```{r echo = FALSE, message = FALSE, warning=FALSE}
# Superimpose the occupancy and the intensity weighted occupancy 
IWO_plot <- Weight_moves_plot + 
   geom_line(aes(y = Occupancy), colour = "grey") + 
   geom_line(aes(y = IWO-mean(IWO)+mean(Occupancy)), colour = "red", alpha = 0.5) + 
   labs(title = "Intensity Weighted Occupancy",
        subtitle = "Raw and IWO overplotted",
        y = "Mean-balanced occupancy") 

# Now also superimpose the lowess smoother with 95% CI 
IWO_smooth_plot <- Weight_moves_plot + 
   geom_line(aes(y = Occupancy), colour = "grey") + 
   geom_line(aes(y = IWO-mean(IWO)+mean(Occupancy)), colour = "red", alpha = 0.5) + 
   geom_smooth(aes(y = IWO-mean(IWO)+mean(Occupancy)), colour = "purple", alpha = 0.5) +
   geom_smooth(aes(y = Occupancy), colour = "grey40", alpha = 0.5) +
   labs(title = "Intensity Weighted Occupancy",
        subtitle = "Raw and IWO occupancy with smoothing",
        y = "Mean-balanced occupancy") 

# arrange these two plots side by side - in a grid with 1 row - for comparison 
grid.arrange(IWO_plot, IWO_smooth_plot, nrow = 1)
```

## Exit Block

The time between a decision to discharge from ICU, and actual ICU, can be thought of as time spent in a safe location. There are, however, financial costs associated with such a prolongation of ICU stay, such as staff, or the need to provide more standing resources such as monitors^[
It is not always possible to reduce the per-patient staff ratios if the number of exit blocked patients are not an exact multiple of the lower staffing ratio.
]. Processes of care and planning may also be disrupted due to ICU exit block, due to lack of home team access or more subtle lack of engagement. There may also be more direct interference with processes of care if patients are transferred to other locations where other systems are used. This is exacerbated if the patients are transferred during periods of lower staffing intensity, or where many patients are discharged from ICU at once to provide for urgent admissions. These possibilities are discussed below.

```{r exit.block, echo=FALSE, message = FALSE, warning=FALSE}
exit_block <- APD %>% 
   group_by("Elective" = PlannedAdm) %>% 
   summarise("n" = n(),
             "Total days" = sum(time_length(ICU_DIS_DTM - ICU_DIS_DEC_DTM, unit = "day"), na.rm=T),
             "Average hours" = mean(time_length(ICU_DIS_DTM - ICU_DIS_DEC_DTM, unit = "hour"), na.rm=T),
             "1st quartile" = quantile(time_length(ICU_DIS_DTM - ICU_DIS_DEC_DTM, unit = "hour"), na.rm=T, probs=0.25),
             "3rd quartile" = quantile(time_length(ICU_DIS_DTM - ICU_DIS_DEC_DTM, unit = "hour"), na.rm=T, probs=0.75)
                ) 

kable(exit_block)
```

We assume an ICU bed day costs \$5320 more than a ward bed. A naive cost reduction estimate is \$`r sum(time_length(APD$ICU_DIS_DTM - APD$ICU_DIS_DEC_DTM, unit = "day"), na.rm=T)*5330` over this period, if all exit block were removed. Take a quick glance at the margin plot to check if the figures for discharge delay make sense.

More sensibly, if we prevented only the upper 12.5\% of the exit blocked patients, those with the longest time of exit block, then \$`r sum(APD$ExitBlockHrs[APD$ExitBlockHrs>(quantile(APD$ExitBlockHrs, 0.875, na.rm=T))], na.rm=T)*5330/24` would be saved for the healthcare system during this period alone.

```{r check.exit.block, message = FALSE, fig.margin = TRUE, fig.cap = "Check the exit block numbers make sense.", fig.width=3.5, fig.height=3.5, cache=TRUE, message=FALSE, warning=FALSE, echo = FALSE}
ggplot(data=APD, aes(PatientId, ExitBlockHrs/24)) + geom_point() + labs(x = NULL, y="DC delay in days")
```

## Discharge after hours 

Of the `r APD %>% filter(ICUOutcome != "Died") %>% summarise(n())` patients who left ICU alive, `r APD %>% filter(ICUOutcome != "Died" & (hour(ICU_DIS_DTM)>18 | hour(ICU_DIS_DTM)<6)) %>% summarise(n())` were discharged out of hours and a further `r APD %>% filter(ICUOutcome != "Died" & day(ICU_DIS_DTM) %in% c(1,7)) %>% summarise(n())` were discharged on a weekend. The table below compares some of their characteristics. 

```{r readmission.sums, echo=FALSE, message = FALSE}

readmitno <- APD %>% filter(ReadmissionEpisode >0) %>% select(PatientId) %>% nrow()

readmitpats <- APD %>% filter(ReadmissionEpisode >0) %>% select(PatientId) %>% n_distinct()

readmit72hno <- APD %>% filter(ReadmissionEpisode > 0 & ReadmissionLagHours<=72) %>% nrow()

readmit72hpats <- APD %>% filter(ReadmissionEpisode > 0 & ReadmissionLagHours<=72) %>% n_distinct()

readmit72mort <- round(mean(APD$HospitalOutcome[APD$ReadmissionEpisode >0 & APD$ReadmissionLagHours<=72]=="Died", na.rm=T), 3)

readmitmort <- round(mean(APD$HospitalOutcome[APD$ReadmissionEpisode >0]=="Died", na.rm=T), 3)

firstmort <- round(mean(APD$HospitalOutcome[APD$ReadmissionEpisode ==0]=="Died", na.rm=T), 3)

OOH_discharges <- APD %>%
   filter(ICUOutcome !="Died") %>% 
   group_by(OOHDC) %>% 
   summarise(
      "Totals" = n(),
      "Admission ANZROD" = round(mean(ANZROD, na.rm=T), 3),
      "ICU LOS" = round(mean(ICULOS, na.rm=T)/24, 2),
      "Hospital mortality" = round(mean(HospitalOutcome=="Died", na.rm=T), 3),
      "Post ICU LOS" = round(mean(interval(ICU_DIS_DTM, HOSP_DIS_DTM)/86400, na.rm=T),3)
      ) %>% kable()

OOH_discharges
```

## Readmission rate 

In sympathy with the language of patient journeys, a readmission can be viewed as a backward step. The expectation may be that on discharge from ICU, patients are either a path to wellness or death and that unexpected divagations from that path are unwelcome. It is not possible to determine from the data whether any of these unexpected movements are avoidable. In this period, `r readmitpats` patients were readmitted, accounting for `r readmitno` readmissions in total, `r readmit72hno` of those being within 72 hours of ICU discharge. The hospital mortality rate was `r readmitmort`, or `r readmit72mort` in those readmitted within 72 hours, compared to `r firstmort` for first admissions. 

## Data quality 

There are many ways to describe missingness. This is a big topic but the basic difference is between entries that are missing at random, and those where the missingness itself represents a process. It's difficult to tell from a single dataset which is happening, and topic knowledge is needed. We get a report on the data quality of the essential items, but not on the whole dataset. Here is a snapshot of the proportion of missingness in each variable.

```{r Missing.plot, fig.margin = TRUE, fig.cap = "APD variables ordered by proportion missing over the period.", message=FALSE, warning=FALSE, echo = FALSE}
missvarsplot <- naniar::gg_miss_var(APD,  show_pct = TRUE) +
   theme(axis.text.y=element_blank())

misscasesplot <- naniar::gg_miss_case(APD, show_pct = TRUE)

grid.arrange(missvarsplot, misscasesplot, nrow = 2)
```

The table below gives further detail on the variables with the greatest proportion of missing data. Many of these are simply not useful (for example the fields for CABG and Transfers, which have been removed by the code). Many are extremely useful, but missed for various reasons.

```{r message = FALSE}
APD %>% select(-EthnicityOther, -CABGGraft, -CABGRedo, -PT_identifier,
               -ICU_identifier, -HOSP_identifier, -AMIThromboTher, -ReadmissionLagHours,
               -TransferredFrom, -TransferredTo) %>% 
   naniar::miss_var_summary(n=20)
```
In exploring the reasons for missing observations, some plots can help. These are not necessarily what is needed in a monthly report, but the patterns of missingness, as a signal of the underlying mechanisms by which observations are missed, can be very valuable in examining our own practice.

# ANZROD and outcome 

We want to know if all of this activity has been to our stated purposes. The chief among these purposes is saving lives. First is to determine how the severity relates to outcomes. 

To begin that story: this slightly busy plot repeats the above analysis, and codes the admissions by outcome. Those who survived are a shade of blue, and those who died are a shade of red. The lowest risk of death is displayed in the deepest red, to emphasise those patients who went on to die against the prediction of the algorithm, who will appear as **short bright bars**. These outcomes are [distressing](https://www.beyondblue.org.au/) and often raise the question of whether they could have been prevented. 

```{r echo = FALSE, message = FALSE, warning=FALSE, fig.cap = "Short bright red bars suggest an ICU death in a patient with a low predicted mortality and brief ICU stay before death."}
# This overlaps all the patient stays, colouring by outcome with density of colour by ANZROD 
occupancy_intensity +
   geom_rect(data = APD %>% filter(ICUOutcome!='Died'), mapping = aes(xmin = ICU_ADM_DTM, xmax = ICU_DIS_DTM, ymin = 0, ymax = ANZROD, fill = ANZROD, alpha = 0.05)) + 
   scale_fill_continuous(low="azure", high="blue2") +
   labs(title = "Intensity weighted bed occupancy", 
        subtitle = "Stratified by outcome of ICU admission",
        fill = "ANZROD if survived") +    
   new_scale_fill() + 
   geom_rect(data = APD %>% filter(ICUOutcome=='Died'), mapping = aes(xmin = ICU_ADM_DTM, xmax = ICU_DIS_DTM, ymin = 0, ymax = ANZROD, fill = ANZROD, alpha = 0.05)) + 
   scale_fill_continuous(low="red", high="hotpink2") +
   labs(fill = "ANZROD if died") +
   theme(axis.text.y = element_blank()) +
   guides(guide_legend(alpha = element_blank()))

roddead <- round(median(APD$ANZROD[APD$ICUOutcome=="Died"], na.rm=T), 3)

rodlived <- round(median(APD$ANZROD[APD$ICUOutcome!="Died"], na.rm=T), 3)

```

## The bigger picture 

We often only review the predicted risk of death for those who died, because that is the structure of our quality reporting. Over this period, the table below shows the ANZROD-predicted risk of death for those who died was `r roddead`. As a corrective, it's important also to review the predicted risk of death in survivors. Across this period the median predicted risk of death in those who actually survived was `r rodlived`, as shown in the table below. 

```{r ANZROD_by_outcome-table, fig.margin = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
ANZROD_by_outcome <- kable(APD %>% 
   group_by(ICUOutcome) %>% 
   summarise("Mean ANZROD" = paste(
      round(median(ANZROD, na.rm=T), 2),
      "+/-", 
      round(IQR(ANZROD, na.rm=T)/2, 2), 
      sep = " ")
      ),
   caption = "Algorithmically predicted mortality for each category of ICU outcome during the period. Median and interquartile range are shown. The range is wide where few patients are in that category, or where the mortality predictions for each patient were very divergent."
   )
ANZROD_by_outcome
```

More important again, however, is the outcome in each category of predicted risk. That is listed in the table below (which might have some gaps because of small numbers, depending on the period selected).

```{r ANZROD.quintiles.table, echo=FALSE, message = FALSE}
APD %>% 
   select(ANZROD, ICUOutcome) %>% 
   mutate(RODquintile = ntile(ANZROD, 5)) %>% 
   group_by(RODquintile) %>% 
   summarise(
      Patients = n(),
      `Count of outcomes` = sum(ICUOutcome=="Died", na.rm=T), 
      `Observed mortality` = round(mean(ICUOutcome=="Died", na.rm=T), 3)) %>%
    mutate("Calculated ROD" = round(quantile(APD$ANZROD, probs = seq(0, 1, 0.2), na.rm=T), 3)) %>% 
   kable()
                                           
```

## The Standardised Mortality Ratio 

Although we all die, each death of another human marks us, and should mark us. In the M&M we should talk about the specific circumstances of each death, we should mark the lives that end here and never fool ourselves into thinking that our jobs are just another process that can be controlled. 

At the same time, we need to be sure that all our human actions and our care are directed in the right way. We need to be able to tell ourselves, our families and the public that we are worth the trust that is placed in us. And if you care about that sort of thing, we need to benchmark and report to managers. 

The predictions for risk of death are imprecise, but are well calibrated (meaning that on average, a higher ANZROD, for example, means a higher risk of death no matter where you are). This means that ANZROD can be used to compare the risk of hospital death across ICUs as long as some basic elements such as case mix, patient selection, staffing, and discharge practices are the same across those units. 

Because we can compare the risk of death, we can then compare the risk of death to the actual deaths. This generates the indirectly standardised mortality ratio or SMR
^[Indirect standardisation, as here, uses a model of some kind. Direct standardisation uses the actual deaths in the relevant strata of each population, usually defined by age and sex. Direct standardisation is better when talking about population health, as it doesn't miss out on those who failed to come to ICU. For ICU processes and outcomes, indirect standardisation by SMR allows us to take account simultaneously of all the variables in the ANZROD, and so is better.]
. 

The SMR by month in this period is listed below. Remember a value of **less than 1** for SMR means that fewer people died than expected. Remember as well that ANZROD predicts hospital death, and so the numbers are higher than for ICU deaths. The ratio of predicted to observed deaths across the entire period was `r round(sum(APD$HospitalOutcome=="Died", na.rm=T)/sum(APD$ANZROD, na.rm=T), 3)`.

```{r SMR, echo=FALSE, message = FALSE}
SMRmonth <- APD %>% 
   select(ANZROD, HospitalOutcome, ICU_ADM_DTM) %>% 
   mutate(Month = month(ICU_ADM_DTM)) %>% 
   group_by(Month) %>% 
   summarise(
      Patients = n(),
      Predicted = round(mean(ANZROD, na.rm=T), 3), 
      Observed = round(mean(HospitalOutcome=="Died", na.rm=T),3),
      SMR = round(Observed/Predicted, 3)
      ) 

SMRmonth %>% kable(caption = "Observed versus predicted mortality, and resulting SMR by month")
```

```{r SMRplot, echo=FALSE, message = FALSE}
# Next I'll overplot the admissions and SMR with error bars. 

#SMRplot <- ggplot(data = SMRmonth, aes(x = Month, y = SMR))+
#   geom_smooth() +
#   geom_bar(Patients) + Stat_identity()
#SMRplot

#admission_rate + geom_histogram(fill = "pink", alpha = 0.9, #show.legend = FALSE)+
#   labs(title = "Admission rate over the period",
#        x = "Admission date",
#        y = "Admissions per week")
```

```{r}
ICUtrache <- APD %>% filter(! APACHEDiagnosis %in% c(202, 208, 1303) & Trachestomy==TRUE) %>% count()
```

# Additional questions{#question}

The code below has some skeleton elements (such as the notes on how to reproduce the analysis), and some elements that are intended directly to answer the question for today. 

## Subsets of admissions 

The count of certain admission diagnoses or clinical features can be shown in the context of all admissions. Here, cardiovascular admissions are displayed. The first code returns the diagnostic category numbers for any diagnoses admitted during the period. Use these to fill in the second bit of code with the diagnostic categories you're interested in.

```{r APACHE.categories, fig.margin = TRUE, message = FALSE, warning = FALSE}
#retrieves the categories present in the dataset 
levels(as.factor(APD$APACHECategory))
```

Use these subsets to create the graphs below (or tables, from almost the same code).

```{r Diagnosis_admissions, echo = FALSE, message = FALSE, warning=FALSE}
#plots stacked bar charts for the total and group specific admissions. Use the numbers in the concatenation brackets (the ones inside `c(...)`) to mark the ones you want displayed. 
admission_trauma <- ggplot(data = APD, mapping = aes(x = ICU_ADM_DTM,  fill = (APACHEDiagnosis %in% c(601:605, 1601:1605)))) + 
   geom_histogram() +
   labs(y = "Admissions per week",
        x = "Admission date",
        fill = "Trauma") + 
   scale_fill_brewer(palette = "Pastel2")

admission_cv <- ggplot(data = APD, mapping = aes(x = ICU_ADM_DTM,  fill = (APACHEDiagnosis %in% c(101:111, 1202:1213)))) + 
   geom_histogram() +
   labs(y = "Admissions per week",
        x = "Admission date",
        fill = "CVS") + 
   scale_fill_brewer(palette = "Pastel2")

admission_headtrauma <- ggplot(data = APD, mapping = aes(x = ICU_ADM_DTM,  fill = (APACHEDiagnosis %in% c(601)))) + 
   geom_histogram() +
   labs(y = "Admissions per week",
        x = "Admission date",
        fill = "TBI") + 
   scale_fill_brewer(palette = "Pastel2") 

grid.arrange(admission_cv, admission_headtrauma, nrow=1)
```


The full admission numbers over the period are also displayed in the table below, with actual ICU mortality and predicted mortality by the ANZROD algorithm. 

```{r echo = FALSE, message = FALSE, warning=FALSE}
# This uses the mortality calculation from the setup chunk at the top of the document, use the commented out section to select only those where there is a valid comparison to be made, ie those with any actual deaths.

#APDmortality %>% filter(`Observed Mortality` > 0.001) %>%  kable()

APDmortality %>% arrange(desc(Admissions)) %>% kable()
```

## ECMO need

**How many patients are likely to benefit from ECMO retrieval?** 

```{r echo = FALSE, message = FALSE, warning=FALSE}
# This uses the calculated PaO2:FiO2 ratio (as APD$PF) for selection 
VV_candidates <- APD %>% 
   filter(PF<=150 & 
             Intubated == TRUE & 
             (ICUOutcome=="Died" | (HospitalOutcome=="Died" & ICULOS/24>6) | HOSLOS/24>30) & 
             AgeICU <= 75 &
             TreatmentGoals=="Full Active Treatment") %>%
   filter(across(ImmuneDisease:Leukaemia, ~ .x == FALSE))

VA_candidates <- APD %>% 
    filter(APACHEDiagnosis %in% 101:111 &
              (ICUOutcome== c("Died","OtherAcuteHospitalIcu") | HospitalOutcome=="Died") &
              AgeICU <= 75 &
             TreatmentGoals=="Full Active Treatment") %>% 
    filter(across(ImmuneDisease:Leukaemia, ~ .x == FALSE))

VV_strict <- VV_candidates %>%  
    filter(PF<=100 & AgeICU<=65) %>% 
    filter(across(ChronicRespiratory:Leukaemia, ~ .x == FALSE))

VA_strict <- VA_candidates %>% 
    filter(AgeICU <= 65)%>% 
    filter(across(ChronicRespiratory:Leukaemia, ~ .x == FALSE))

VVbasemort <- sum(VV_candidates$ICUOutcome=="Died") / tally(VV_candidates)

VVbenefit <- round(count(VV_candidates)*(1-(VVbasemort-0.15)),1)

costperVVbenefit <- 20000*tally(VV_candidates)/VVbenefit
```

This question can't be directly answered with retrospective data. The top line answer is that between `r tally(VV_strict)` and `r tally(VV_candidates)` patients met criteria for VV ECMO, and between `r tally(VA_strict)` and `r tally(VA_candidates)` patients for VA ECMO per year. 

For the `r count(VV_candidates)` patients who met VV ECMO oxygenation criteria, `r sum(VV_candidates$ICUOutcome=="Died")` or `r round(VVbasemort*100, 1)`% died. An estimated benefit of 15% absolute mortality reduction suggests that at the very least `r VVbenefit` extra Territorians would be alive at each year's end with access to ECMO facilities. At a cost of $20,000 per VV ECMO run this equates to `r as.integer(costperVVbenefit)` dollars per survivor, well within common cost:benefit criteria. 

There may be patients who survived but with impairments that could have been prevented with more aggressively lung protective ventilation. 

There may be patients who were considered unsalvageable and were not even referred for consideration of ICU. There may be patients who are listed with one top line diagnosis but who also had a condition amenable to mechanical circulatory or ventilatory support. Registry data is unreliable on the inclusions and exclusions which pertain at the moment of considering referral. 

As a starting point, we chose set of criteria which reflect the experience of ELSO sites, before further limiting the criteria to those patients with a high predicted benefit _if their registry entries reflect their determinative parameters_. The latter number places a lower limit on the possible referrals: 

    1 For VV ECMO:
        - Worst P:F ratio <100 mmHg / atm while invasively ventilated 
        - Died in ICU OR long stay >7 days
        - No chronic conditions preventing VV ECMO treatment, and age <=65
    2 For VA ECMO: 
        - Cardiovascular admission or cardiac arrest 
        - Inotropes required and highest Systolic BP<90 
        - Died or transferred to another ICU 
        - No chronic conditions preventing VV ECMO treatment, and age <=65

This represents the best available evidence on which to base the decision. 

## Missed Sepsis 
Missed sepsis is a politically important item. A pathway has been developed for this. Identification of cases may follow the same history seen in VAP and other politically sensitive system diagnoses. One simple KPI relates to patients admitted to ICU with a diagnosis of sepsis from the ward, within 48h of admission to hospital.

```{r missed-sepsis-numbers, echo = FALSE, message = FALSE, warning=FALSE}

missed_sepsis_ROD <- APD %>%
   ggplot(aes(y=ANZROD, x=missed_sepsis))+
   geom_violin()

missed_sepsis_ROD

missed_sepsis_table <- APD %>% group_by(missed_sepsis) %>%
   summarise(n = n(),
             ANZROD = round(mean(ANZROD, na.rm=T), 2),
             Mortality = round(mean(DiedICU=="Died", na.rm=T), 3),
             "ICU LOS" = round(mean(ICULOS, na.rm=T)/24, 2),
             Age = round(median(AgeICU), 2))
             
kable(missed_sepsis_table)
yearsofblock <- tibble(year = seq(2008, 2020, 1), Admissions = c(776, 808, 917, 901, 950, 956, 915, 939, 958, 1158, 1104, 1121, 1060), `Exit Block` = c(180, 329, 351, 459, 439, 402, 319, 746, 737, 624, 1184, 889, 771))

ggplot(data=yearsofblock) +
    geom_line(aes(x = year, y = `Exit Block`), colour = "red") +
    geom_line(aes(x=year, y=Admissions), colour = "black") +
    geom_smooth(method = "lm", aes(x = year, y = `Exit Block`), colour = "red") + 
    geom_smooth(method = "lm", aes(x=year, y=Admissions), colour = "black") +
    labs(title = "Admissions (black) vs exit block days", y = "Exit Block in days / Admissions per year")

#anova(lm(yearsofblock$Admissions ~ yearsofblock$year), lm(yearsofblock$`Exit Block` ~ yearsofblock$year))

#whys <- tibble(
#    causes = c("Total Beds", "Ward delays in DC", "Priority Allocation", "Ward skill shortage", "ICU staffing inflexible", "Communication about exit", "Infection Isolation", "ICU delays plans", "Top-down culture"),
#    votes = c(24, 15, 12, 9, 7, 6, 5, 4, 3)
#    )

#exitpareto1 <- ggplot(data=whys, aes(x = reorder(causes, desc(votes)))) +
#    geom_line(mapping = aes(x=reorder(causes, desc(votes)), y = cumsum(desc(-votes)), group = 1)) + 
#    geom_col(aes(x=reorder(causes, desc(votes)), y = votes, fill = causes)) +
#    scale_fill_brewer() +
#    labs(title = "Pareto chart of opinions on ICU exit block", x = "causes", y = "proportion") +
#    theme(axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=0.75))

#exitpareto1

#library(qcc)
#pareto.chart(whys$votes)
```

As summarised in the table, there were `r sum(APD$missed_sepsis==TRUE)` admissions in this category, of `r length(unique(APD$SLK)[APD$missed_sepsis==TRUE])` individuals. The predicted mortality in this sub-cohort was `r round(mean(APD$ANZROD[APD$missed_sepsis==TRUE]), 3)`. The actual mortality in ICU was `r round(mean(APD$ICUOutcome[APD$missed_sepsis==TRUE]=="Died",na.rm=T),3)`, and hospital mortality was `r round(mean(APD$HospitalOutcome[APD$missed_sepsis==TRUE]=="Died",na.rm=T),3)`.

## Demand management for exit block 

Although it may be desirable to improve discharge processes and efficiency, it may be simpler to provide a safe space to care for these patients outside of ICU. Once suggestion relates to lower risk elective surgical patients, grouped in the first case by elective procedure. 

```{r low-risk-numbers, echo = FALSE, message = FALSE, warning=FALSE}
procedureprobs <- APD %>% 
    filter(!is.na(APACHESubcode)) %>% 
    group_by(APACHESubcode) %>% 
    mean(DiedICU, na.rm=T)

lowriskproc <- APD %>% 
   filter(APACHESubcode %in% c(1203.02, 1205.01, 2201.05)) 
surgnointervention <- APD %>% 
   filter(ElectiveSurg == TRUE) %>% 
   filter(nointervention == TRUE)

```

There were `r nrow(lowriskproc)` patients with lower risk procedures, with an average length of stay of `r round(mean(lowriskproc$ICULOS, na.rm=T)/24, 2)` days (and an average exit block of `r round(mean(lowriskproc$ExitBlockHrs, na.rm=T), 2)` hours). `r nrow(filter(lowriskproc, nointervention == TRUE))` of these patients received no ICU level intervention. Including these patients, `r nrow(surgnointervention)` elective surgical admissions involved no ICU level interventions, during their average `r round(mean(surgnointervention$ICULOS, na.rm=T)/24, 2)` days of which `r mean(surgnointervention$ExitBlockHrs, na.rm=T)` hours were exit blocked. It may not be possible to identify these low risk patients beforehand; some of their admission characteristics are summarised below. Identification of such low risk patients may indeed repay some of the required effort, given the estimated \$`r sum(surgnointervention$ICULOS, na.rm=T)/24*5330` these admissions cost.

```{r nointervention.anzrod, echo = FALSE, message = FALSE, warning=FALSE}
surgnointer.table <- APD %>% 
   filter(ElectiveSurg == TRUE) %>% 
   group_by("Free from intervention" = nointervention) %>% 
   summarise(n = n(),
             Age = round(median(AgeICU, na.rm = T), 2),
             ANZROD = round(mean(ANZROD, na.rm = T), 2),
             APACHE = round(mean(AP3score, na.rm = T), 2),
             Female = round(mean(Sex=="Female"), 2))

noneed_ANZROD <- APD %>%
   ggplot(aes(y=ANZROD, x=nointervention))+
   geom_violin() +
   labs(x = "Intervention free")

kable(surgnointer.table)
```